From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-28 Pi/paper/discussion.tex
Miscellaneous changes

Coherence theorems are famous problems in category theory, and Mac Lane's coherence
theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems
for higher categorical structures has a long history, see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$~\cite[Theorem 3.5]{ortonDecomposingUnivalenceAxiom2018}.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

>>>>

Coherence theorems are famous problems in category theory, and Mac Lane's coherence
theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems for
higher categorical structures has a long history,
see~\cite{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$~\cite[Theorem 3.5]{ortonDecomposingUnivalenceAxiom2018}.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse. Univalent
typoids~\cite{petrakisUnivalentTypoids2019} are another way of axiomatising univalent subuniverses.

% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which provides
an intuition on why certain constructions are hard (or impossible in the general case). There are other, different
approaches to computing with univalence,
in~\cite{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in Cubical
Type Theory\cite{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

### 2021-10-28 Pi/paper/examples.tex
Change size of qiskit figures
Use qiskit svg figures
Miscellaneous changes

For concreteness, we demonstrate two different circuits that implement the following reversible function specification
$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface \verb|qasm|
and the graphical interface in~\cref{fig:qiskit}.

\begin{figure}
  \begin{tabular}{c@{\qquad}c}
    \begin{minipage}[t]{0.42\linewidth}
      \begin{verbatim}
  ccx q[1], q[2], q[0];
  cx  q[1], q[0];
  cx  q[2], q[0];
  \end{verbatim}
      \includegraphics[scale=0.7]{reversibleOr.png}
    \end{minipage}
     &
    \begin{minipage}[t]{0.43\linewidth}
      \begin{verbatim}
  cx  q[1], q[0];
  x   q[1];
  ccx q[1], q[2], q[0];
  x   q[1];
  \end{verbatim}
      \includegraphics[scale=0.6]{reversibleOr2.png}
    \end{minipage}
  \end{tabular}
  \caption{$\mathit{reversibleOr}$ in Qiskit}
  \label{fig:qiskit}
\end{figure}

There is a wealth of manual and algorithmic approaches for producing circuits such as these
two~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the left was manually produced using a standard synthesis
algorithm for reversible circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced using an approach
that analyzes the recursive structure of $\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
more than two inputs):

@@

Although rather trivial, the examples above illustrate the general idea. A more interesting example would be the
classical core of Shor's algorithm which requires a circuit implementing modular exponentiation $f(r) = a^{r} \mod N$
for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a reversible specification of this
function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive disjunction~$\underline{\vee}$ is applied
bitwise. When $h=0$ the second component of the result of $f^R$ is the result of the original modular exponentiation
$f$.
%%
%% the specification of the circuit is relatively straightforward to
%% calculate. Here it is for $a=11$ and $N=15$:
%% \[\begin{array}{rcll}
%% g(r,h) &=& \left\{ \begin{array}{ll}
%%                      (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
%%                      (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
%%                      (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
%%                      (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
%%                                 \end{array}\right.
%% \end{array}\]
%%
However, as explained in standard accounts of the algorithm (e.g., the
Qiskit implementation), producing an efficient modular exponentiation circuit
from this specification is not straightforward and is actually the bottleneck in
Shor’s algorithm. Typical derivations of the circuit start from elementary
gates, build a circuit for reversible disjunction (like the two circuits above),
reversible conjunction, a circuit for a half-adder, a circuit for computing the
carry, progressing to a circuit for modular addition, which is used to build a
circuit for modular multiplication, and then finally a circuit for modular
exponentiation taking care at each step to avoid the exponential blowup (e.g.,
by implementing exponentiation by squaring instead of repeated
multiplication)~\cite{shorefficient}.

>>>>

For concreteness, we demonstrate two different circuits that implement the following reversible function specification
$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface \verb|qasm|
and the graphical interface, in~\cref{fig:qiskit}.

\begin{figure}
  \begin{tabular}{c@{\qquad}c}
    \begin{minipage}[t]{0.45\linewidth}
      \begin{BVerbatim}[fontsize=\small]
        ccx q[1], q[2], q[0];
        cx  q[1], q[0];
        cx  q[2], q[0];
  \end{BVerbatim}
      \inkscape[0.5]{qis1}
    \end{minipage}
     &
    \begin{minipage}[t]{0.45\linewidth}
      \begin{BVerbatim}[fontsize=\small]
        cx  q[1], q[0];
        x   q[1];
        ccx q[1], q[2], q[0];
        x   q[1];
  \end{BVerbatim}
      \inkscape[0.5]{qis2}
    \end{minipage}
  \end{tabular}
  \caption{$\mathit{reversibleOr}$ in Qiskit}
  \label{fig:qiskit}
\end{figure}

There is a wealth of manual and algorithmic approaches for producing circuits such as these
two~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the left was manually produced using a standard synthesis
algorithm for reversible circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced using an approach
that analyzes the recursive structure of $\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
more than two inputs).

@@

Although rather trivial, the examples above illustrate the general idea. A more interesting example would be the
classical core of Shor's algorithm which requires a circuit implementing modular exponentiation $f(r) =
a^{r}~\mathrm{mod}~N$ for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a reversible
specification of this function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive
disjunction~$\underline{\vee}$ is applied bitwise. When $h=0$ the second component of the result of $f^R$ is the result
of the original modular exponentiation $f$.
%%
%% the specification of the circuit is relatively straightforward to
%% calculate. Here it is for $a=11$ and $N=15$:
%% \[\begin{array}{rcll}
%% g(r,h) &=& \left\{ \begin{array}{ll}
%%                      (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
%%                      (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
%%                      (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
%%                      (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
%%                                 \end{array}\right.
%% \end{array}\]
%%
However, as explained in standard accounts of the algorithm (e.g., the
Qiskit implementation), producing an efficient modular exponentiation circuit
from this specification is not straightforward and is actually the bottleneck in
Shor’s algorithm. Typical derivations of the circuit start from elementary
gates, build a circuit for reversible disjunction (like the two circuits above),
reversible conjunction, a circuit for a half-adder, a circuit for computing the
carry, progressing to a circuit for modular addition, which is used to build a
circuit for modular multiplication, and then finally a circuit for modular
exponentiation taking care at each step to avoid the exponential blowup (e.g.,
by implementing exponentiation by squaring instead of repeated
multiplication)~\cite{shorefficient}.

### 2021-10-28 Pi/paper/ufin.tex
Miscellaneous changes

In categorical language, the setting for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA2010}.

The semantics interprets types of $\PiLang$ as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of 1-combinators
related by a 2-combinator, their interpretations in~$\BFin$ are equal.  The groupoid $\BFin$ is strict, since the
collection of isomorphisms is a set, that is, a discrete category. There is no explicit witness for the equality of two
isomorphisms, since we can decide by evaluating two bijections whether they are equal. To be able to establish
completeness for $\PiLang$, we want a witness for this equality, so that we can quote back to the syntax and produce a
2-combinator witnessing the equality of the corresponding 1-combinators.

@@

We observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells.
We work in HoTT (Univalent Foundations) as it provides a proof-relevant, constructive metatheory to get a handle
on these equalities and provides a rich internal language for describing weak groupoids, using the ``types are weak
$\infty$-groupoids'' correspondence.

@@

The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is the total space of $P$. The first projection
${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a fibration, that
is, there is a lifting operation (\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in the
total space.

@@

Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes a set $A$ and a binary relation $R : {A \to A \to
  \UU}$. It has a mapping of points $\quotinc : A \to \quot{A}{R}$, and a constructor that adds paths between related
pairs of elements $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$ (see~\cref{def:set-quot}).  We
recall that the quotient is \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff
$(q(x) \id_{\quot{A}{R}} q(y))$.

@@

A type is \emph{contractible} (-2-type) if it has a unique element, that is, there is a center of contraction and every
other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that is, it
has at most one inhabitant. Iterating this, we can define sets or 0-types (whose equality types are propositions) and
1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.
\begin{gather*}
  \begin{aligned}
    \isContr{A} & \defeq \dsum*{x:A}{\dfun*{y:A}{y \id x}} \\
    \isProp{A}  & \defeq \dfun*{x,y:A}{\isContr{x \id y}}
  \end{aligned}
  \qquad
  \begin{aligned}
    \isSet{A} & \defeq \dfun*{x,y:A}{\isProp{x \id y}} \\
    \isGpd{A} & \defeq \dfun*{x,y:A}{\isSet{x \id y}}
  \end{aligned}
\end{gather*}

\noindent
Given a function $f : A \to B$, there is an associated fibration over $B$, given by the homotopy fiber. The image of $f$
is defined to be the (-1)-truncation of its fiber.
\begin{gather*}
  \fib_{f}(b) \defeq \dsum*{a:A}{f(a) \id_{B} b}
  \qquad
  \im{f} \defeq \dsum*{b:B}{\Trunc[-1]{\fib_{f}(b)}}
\end{gather*}

\begin{proposition}
  \label{prop:three-chars-equiv}
  The following propositions are equivalent.
  \begin{enumerate}
    \item $f : A \to B$ is an equivalence.
    \item $f$ has a left and right inverse.
    \item $f$ has contractible fibers.
  \end{enumerate}
\end{proposition}

@@

\begin{propositionrep}[Univalent Subuniverse]
  \label{prop:univsub}
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. Given such a predicate $P$, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is
  univalent and generates a univalent subuniverse ${\UU_{P} \defeq (\dsum*{X:\UU}{P(X)}, \fst)}$.~\footnote{Univalent
    typoids~\cite{petrakisUnivalentTypoids2019} are a different presentation of univalent subuniverses.}
\end{propositionrep}

@@

\begin{definition}[$\Fin$]
  \label{def:finite-set}
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
  equivalently in two different ways,
  \begin{gather*}
    \Fin[n] \defeq \dsum*{k:\Nat}{k < n}
    \qquad\qquad \text{or} \qquad\qquad
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a set, and we use both definitions interchangeably.
\end{definition}

\begin{definition}[$\isFin$]
  \label{def:isfin}
  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n : \Nat$.
  \[
    \isFin[X] \defeq \dsum*{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
  Note that the natural number $n$ need not be truncated, as justified below.
\end{definition}

@@

Since $\isFin$ is a predicate on the universe $\UU$, we easily get our univalent subuniverse $\UFin$.

@@

This definition of the groupoid of finite types has also been considered
in~\cite{yorgeyCombinatorialSpeciesLabelled2014}. While $\UFin$ has \emph{all} the finite types, we are also interested
in constructing a subuniverse of finite types of a specified cardinality. To do so, we will start with the subuniverse
$\BAut[T]$, for any type $T : \UU$. \footnote{Characterisations of univalent fibrations using the $\BAut$ construction
  have been studied by~\citet{christensenCharacterizationUnivalentFibrations2015}.}

@@

Using $\BAut$, we can talk about types that are equivalent to a finite set of specified cardinality, for example, the
subuniverse of 2-element sets is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give the denotational semantics for a 1-bit reversible
programming language~\cite{caretteReversibleProgramsUnivalent2018}.

@@

We have shown that loops in $\UFin$ exactly encode the automorphism group $\Aut[\Fin[n]]$ for every~$n$. This is a
general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, internally in HoTT.
This technique also allows defining higher groups~\cite{buchholtzHigherGroupsHomotopy2018}. The loopspace of a pointed
type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element, path composition for the
group multiplication, and path inverse for the group inverse operation. The group axioms are given by the higher paths
corresponding to groupoid laws.

@@

\begin{proposition}
  For any $n, m : \Nat$, and for any types $X, Y, Z$,
  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot                  \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\
      \\
      \bot \sqcup X          & \eqv X                     \\
      X \sqcup \bot          & \eqv X                     \\
      (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y             & \eqv Y \sqcup X            \\
      X \times \bot          & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top                             \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\
      \\
      \top \times X          & \eqv X                                \\
      X \times \top          & \eqv X                                \\
      (X \times Y) \times Z  & \eqv X \times (Y \times Z)            \\
      X \times Y             & \eqv Y \times X                       \\
      X \times (Y \sqcup Z)  & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}

>>>>

In categorical language, the target for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA2010}.

The semantics interprets types of $\PiLang$ as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of
1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal. $\BFin$ is a 1-groupoid, where the
collection of isomorphisms is a set. We can evaluate two bijections on their inputs to decide if they are equal, and
there is no higher cell witnessing the equality of two isomorphisms. To be able to establish completeness for $\PiLang$,
we want a witness for this equality, so that we can quote back to the syntax and produce a 2-combinator witnessing the
equality of the corresponding 1-combinator.

@@

We observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible
cells. We work in HoTT/UF as it provides a proof-relevant, constructive metatheory to get a handle on these equalities
and provides a rich internal language for describing weak groupoids, using the ``types are weak $\infty$-groupoids''
correspondence.

@@

The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is the total space of $P$. The first projection
${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a fibration, that
is, there is a lifting operation (see~\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in
the total space.

@@

Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes a set $A$ and a binary relation
$R : {A \to A \to \UU}$. It has a mapping of points $\quotinc : A \to \quot{A}{R}$, and a constructor that adds paths
between related pairs of elements $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$
(see~\cref{def:set-quot}). We recall that if $R$ is a prop-valued equivalence relation, then the quotient is
\emph{effective}, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}} q(y))$.

@@

A type is \emph{contractible} (a~-2-type) if it has a unique element, that is, there is a center of contraction and
every other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that
is, it has at most one inhabitant. Iterating this, we can define~\emph{sets} or 0-types (whose equality types are
propositions) and~\emph{1-groupoids} or 1-types (whose equality types are sets), and similarly, higher homotopy
$n$-types.
\begin{gather*}
  \begin{aligned}
    \isContr{A} & \defeq \dsum*{x:A}{\dfun*{y:A}{y \id x}} \\
    \isProp{A}  & \defeq \dfun*{x,y:A}{\isContr{x \id y}}
  \end{aligned}
  \qquad
  \begin{aligned}
    \isSet{A} & \defeq \dfun*{x,y:A}{\isProp{x \id y}} \\
    \isGpd{A} & \defeq \dfun*{x,y:A}{\isSet{x \id y}}
  \end{aligned}
\end{gather*}

\noindent
Given a function $f : A \to B$, there is an associated fibration over $B$, given by its homotopy fiber. The image of $f$
is defined to be the (-1)-truncation of its fiber.
\begin{gather*}
  \fib_{f}(b) \defeq \dsum*{a:A}{f(a) \id_{B} b}
  \qquad
  \im{f} \defeq \dsum*{b:B}{\Trunc[-1]{\fib_{f}(b)}}
\end{gather*}

\noindent
Finally, we recall a few characterisations of equivalences of homotopy types.
\begin{proposition}
  \label{prop:three-chars-equiv}
  The following are equivalent.
  \begin{enumerate}
    \item $f : A \to B$ is an equivalence.
    \item $f$ has a left and right inverse.
    \item $f$ has contractible fibers.
  \end{enumerate}
\end{proposition}

@@

\begin{propositionrep}[Univalent Subuniverse]
  \label{prop:univsub}
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. Given such a predicate $P$, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is
  univalent and generates a univalent subuniverse ${\UU_{P} \defeq (\dsum*{X:\UU}{P(X)}, \fst)}$.
\end{propositionrep}

@@

\begin{definition}[$\Fin$]
  \label{def:finite-set}
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
  equivalently in two different ways,
  \begin{gather*}
    \Fin[n] \defeq \dsum*{k:\Nat}{k < n}
    \qquad\qquad \text{or} \qquad\qquad
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
\end{definition}

\noindent
Note that $\Fin[n]$ is a set, and we use both definitions interchangeably.

\begin{definition}[$\isFin$]
  \label{def:isfin}
  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n : \Nat$.
  \[
    \isFin[X] \defeq \dsum*{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
\end{definition}

\noindent
Note that the natural number $n$ need not be truncated, as justified below.

@@

\noindent
Since $\isFin$ is a predicate on the universe $\UU$, we easily get our univalent subuniverse $\UFin$. This definition of
the groupoid of finite types has also been considered in~\cite{yorgeyCombinatorialSpeciesLabelled2014}.

@@

While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$, for any type $T : \UU$.
Characterisations of univalent fibrations using the $\BAut$ construction have also been studied
by~\citet{christensenCharacterizationUnivalentFibrations2015}.

@@

Using $\BAut$, we can talk about types that are equivalent to a finite set of specified cardinality, for example, the
subuniverse of 2-element sets is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give sound and complete denotational semantics to a 1-bit
fragment of $\PiLang$~\cite*{caretteReversibleProgramsUnivalent2018}.

@@

We have shown that loops in $\UFin$ exactly encode the automorphism group $\Aut[\Fin[n]]$ for every~$n$. This is a
general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, that is, a
0-connected type in HoTT. This technique also allows defining higher groups~\cite{buchholtzHigherGroupsHomotopy2018}.
The loopspace of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element,
path composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are
given by the higher paths corresponding to groupoid laws.

@@

\begin{proposition}
  For any $n, m : \Nat$, and for any types $X, Y, Z$,
  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot                  \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\[1.5ex]
      \bot \sqcup X          & \eqv X                     \\
      X \sqcup \bot          & \eqv X                     \\
      (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y             & \eqv Y \sqcup X            \\
      X \times \bot          & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top                             \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\[1.5ex]
      \top \times X          & \eqv X                                \\
      X \times \top          & \eqv X                                \\
      (X \times Y) \times Z  & \eqv X \times (Y \times Z)            \\
      X \times Y             & \eqv Y \times X                       \\
      X \times (Y \sqcup Z)  & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}
